LCS Tool : A Computational Platform for Lagrangian Coherent Structures

### 2014-09-05 main.tex
Move periods outside parentheses
Conflicts:
	main.tex

The right Cauchy-Green strain tensor associated with the flow map is defined as
\begin{equation}
C_{t_0}^t(x_0) \equiv \left[\nabla F_{t_0}^t(x_0)\right]^T \nabla F_{t_0}^t(x_0),
\label{eq:CG}
\end{equation}
measuring Lagrangian strain in the velocity field. This tensor is symmetric and positive definite\parencite{truesdell04}. We label the eigenvalues and eigenvectors of $C_{t_0}^t(x_0)$ as follows:
\[
C_{t_0}^t \xi_i = \lambda_i \xi_i, \qquad 0 < \lambda_1 \leq \lambda_2, \quad i = 1,2;
\]
\begin{equation}
\left|\xi_i\right| = 1, \quad \xi_2 = \Omega \xi_1, \quad \Omega = \left(
\begin{array}{rr}
0 & -1\\
1 & 0
\end{array}
\right).
\label{eq:CG_invariants}
\end{equation}

@@

To compute the normal repulsion of shrinklines, we note that an infinitesimal
normal perturbation to a shrinkline $\gamma$ at its point $r$ grows under the flow map $F_{t_{0}}^{t}$ by a factor of $\lambda_{2}(r)$ in the direction normal to $F_{t_{0}}^{t}(\gamma)$. Similarly, small normal perturbations to a stretchline decay by a factor $\lambda_{1}(r)$ in the direction normal to the evolving stretchline.

@@

The Cauchy-Green strain tensor generally admits point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well defined. This generally occurs at a finite set of isolated points within the computational domain\parencite{delmarcelle94}, and hence the point singularities lie off the computational grid with probability one.

@@

Incompressible flows (i.e. those satisfying $\nabla \cdot v=0$) satisfy the relation $\lambda_1(x_0) \lambda_2(x_0) = 1$ at all points of the computational domain \parencite{arnold78:_mathem}. Incompressibility can be computationally imposed by first calculating $\lambda_2(x_0)$, then setting $\lambda_1(x_0) = 1/\lambda_2(x_0)$ and calculating the strain eigenvectors $\xi_2$ from $\lambda_2$, then $\xi_1$ from the relationship in \cref{eq:CG_invariants}. Experience shows that computing $\lambda_i$ in this order gives higher accuracy than in the reverse order \parencite{farazmand12:_comput_lagran}. This procedure should be applied carefully since by changing $\lambda_1$ but not $\xi_1$, the error in satisfying the eigenvalue equation for the Cauchy-Green strain tensor may increase.

@@

\begin{table}
\begin{enumerate}
\item Define a local maximization distance.
\item Find all points of the main grid that are local maxima of $\lambda_2$ within a circle whose radius is the local  maximization distance.
\item Define a maximum shrinkline length
\item Integrate a shrinkline forward and backward according to \cref{eq:strainline} and using the largest $\lambda_2$ local maximum as the initial position. Integrate until the shrinkline has attained the maximum shrinkline length, or until it has reached the domain boundary.
\item Flag any remaining local maxima of $\lambda_2$ within the maximization distance of the shrinkline as ineligible initial positions for subsequent shrinklines
\item Continue integrating shrinklines using local maxima of $\lambda_2$ as initial positions until no eligible local maxima of $\lambda_2$ remain.
\item Remove all shrinkline segments within elliptic LCSs.
\end{enumerate}
\caption{Algorithm to calculate positions of repelling LCSs at time $t_0$. The algorithm for attracting LCSs is similar.}
\label{t:Hyperbolic LCS algorithm}
\end{table}

@@

The local maximization distance is the number of grid points over which the Cauchy-Green strain tensor is susceptible to numerical noise from integration. This distance is thus expected to be a function of variable time step integration tolerances.

Prior to initiating shrinkline and stretchline integration, a maximum length is defined. Shrinklines and stretchlines are then integrated until reaching the domain boundary or the maximum length. The maximum length may be seen as safety factor to prevent highly sinuous shrinklines and stretchlines leading to inordinately long computations; the maximum length should be set such that most shrinklines and stretchlines stop at the domain boundary.

Once shrinkline and stretchline integration has begun, it may evolve so as to be within the local minimization distance of an existing shrinkline or stretchline. In other words, the local minimization distance affects where shrinklines and stretchlines begin, but not where they finish.

@@

\lstinputlisting[caption={Double gyre script for elliptic LCSs},label=l:double gyre lambda-line LCS]{listings/double_gyre/lambda_lcs.m}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_strain_lcs}
\caption{Double gyre elliptic and shrinkline LCSs}
\label{fig:double_gyre_lambda_strain_lcs}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_stretch_lcs}
\caption{Double gyre elliptic and stretchline LCSs}
\label{fig:double_gyre_lambda_stretch_lcs}
\end{figure}

@@

The integration time is $T=4L_x/U$, a multiple of a the eddy turnover
time in the flow. \Cref{fig:Bickley_jet_lambda_LCS_full} shows elliptic LCSs of the Bickley jet with the FTLE in the background. The figure demonstrates a poor result since the LCSs change shape significantly between eddies. To address this problem we could adjust numerical parameters such as the resolution, but instead we will analyze lambda-line closed orbits of one vortex for a short timespan.

\begin{figure}
\centering
\includegraphics[width=.85\textwidth]{graphics/bickley_jet/lambda_lcs_full}
\caption{Bickley jet elliptic LCSs}
\label{fig:Bickley_jet_lambda_LCS_full}
\end{figure}

\Cref{fig:Bickley jet lambda lcs convergence} shows convergence of the elliptic LCSs. The LCS Tool script used to produce that figure are based on the script \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m! which is given in \cref{lst:Bickley jet hyperbolic_shear_lcs subset}. It is worth pointing out that in order to obtain closed orbits, $\lambda = 0.995$; with $\lambda = 1$ no closed orbits exist.

@@

\begin{figure}[hbt]
\centering
\includegraphics[width=.7\textwidth]{graphics/bickley_jet/lambda_lcs_convergence_400}
\includegraphics[width=.7\textwidth]{graphics/bickley_jet/lambda_lcs_convergence_500}
\includegraphics[width=.7\textwidth]{graphics/bickley_jet/lambda_lcs_convergence_600}
\caption{Bickley jet closed lambda-lines at varying resolutions}
\label{fig:Bickley jet lambda lcs convergence}
\end{figure}

@@

% Elliptic LCSs
% Define Poincare sections; ...
poincareSection = struct('endPosition',{},'numPoints',{},'orbitMaxLength',{});
...
poincareSection(1).endPosition = [3.3,-32.1;3.7,-31.6];
poincareSection(2).endPosition = [1.3,-30.9;1.9,-31.1];

>>>>

The right Cauchy-Green strain tensor associated with the flow map is defined as
\begin{equation}
C_{t_0}^t(x_0) = \left[\nabla F_{t_0}^t(x_0)\right]^T \nabla F_{t_0}^t(x_0),
\label{eq:CG}
\end{equation}
measuring Lagrangian strain in the velocity field. This tensor is symmetric and positive definite\parencite{truesdell04}. We label the eigenvalues and eigenvectors of $C_{t_0}^t(x_0)$ as follows:
\[
C_{t_0}^t \xi_i = \lambda_i \xi_i, \qquad 0 < \lambda_1 \leq \lambda_2, \quad i = 1,2;
\]
\begin{equation}
\left|\xi_i\right| = 1, \quad \xi_2 = \Omega \xi_1, \quad \Omega = \left(
\begin{array}{rr}
0 & -1\\
1 & 0
\end{array}
\right).
\label{eq:CG_invariants}
\end{equation}

@@

To compute the normal repulsion of shrinklines, we note that an infinitesimal
normal perturbation to a shrinkline $\gamma$ at its point $r$ grows under the flow map $F_{t_{0}}^{t}$ by a factor of $\lambda_{2}(r)$ in the direction normal to $F_{t_{0}}^{t}(\gamma).$ Similarly, small normal perturbations to a stretchline decay by a factor $\lambda_{1}(r)$ in the direction normal to the evolving stretchline.

@@

The Cauchy-Green strain tensor generically admits point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined. This generically arises at a finite set of isolated points within the computational domain\parencite{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

Incompressible flows (i.e. those satisfying $\nabla \cdot v=0$) satisfy the relation $\lambda_1(x_0) \lambda_2(x_0) = 1$ at all points of the computational domain \parencite{arnold78:_mathem}. Incompressibility can be computationally imposed by first calculating $\lambda_2(x_0)$, then setting $\lambda_1(x_0) = 1/\lambda_2(x_0)$ and calculating the strain eigenvectors $\xi_2$ from $\lambda_2$, then $\xi_1$ from the relationship in \cref{eq:CG_invariants}. Experience shows that computing $\lambda_i$ in this order gives higher accuracy than in the reverse order \parencite{farazmand12:_comput_lagran}.

@@

\begin{table}
\begin{enumerate}
\item Define a local maximization distance.
\item Find all points of the main grid that are local maxima of $\lambda_2$ within a circle whose radius is the local  maximization distance.
\item Define a maximum shrinkline length
\item Integrate a shrinkline forward and backward according to \cref{eq:strainline} and using the largest $\lambda_2$ local maximum as the initial position. Integrate until the shrinkline has attained the maximum shrinkline length, or until it has reached the domain boundary.
\item Flag any remaining local maxima of $\lambda_2$ within the maximization distance of the shrinkline as ineligible initial positions for subsequent shrinklines
\item Continue integrating shrinklines using local maxima of $\lambda_2$ as initial positions until no eligible local maxima of $\lambda_2$ remain.
\item Remove all shrinkline segments within elliptic LCSs.
\end{enumerate}
\caption{Algorithm to calculate initial positions of repelling LCSs at time $t_0$. The algorithm for attracting LCSs is similar.}
\label{t:Hyperbolic LCS algorithm}
\end{table}

@@

@@

\lstinputlisting[caption={Double gyre script for elliptic lambda-line LCSs},label=l:double gyre lambda-line LCS]{listings/double_gyre/lambda_lcs.m}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_strain_lcs}
\caption{Double gyre lambda-line and shrinkline LCSs}
\label{fig:double_gyre_lambda_strain_lcs}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_stretch_lcs}
\caption{Double gyre lambda-line and stretchline LCSs}
\label{fig:double_gyre_lambda_stretch_lcs}
\end{figure}

@@

The integration time is $T=4L_x/U$, a multiple of a the eddy turnover
time in the flow. \Cref{fig:Bickley_jet_lambda_LCS_full} shows lambda-line LCS of the Bickley jet with the FTLE in the background. The figure demonstrates a poor result since the LCSs change shape significantly between eddies. To address this problem we could adjust numerical parameters such as the resolution, but instead we will analyze lambda-line closed orbits of one vortex for a short timespan.

\begin{figure}
\centering
\includegraphics[width=.85\textwidth]{graphics/bickley_jet/lambda_lcs_full}
\caption{Bickley jet lambda-line LCSs}
\label{fig:Bickley_jet_lambda_LCS_full}
\end{figure}

\Cref{fig:Bickley jet lambda lcs convergence} shows convergence of the lambda-line elliptic LCSs. The LCS Tool script used to produce that figure are based on the script \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m! which is given in \cref{lst:Bickley jet hyperbolic_shear_lcs subset}. It is worth pointing out that in order to obtain closed orbits, $\lambda = 0.995$; with $\lambda = 1$ no closed orbits exist.

@@

\begin{figure}[hbt]
\centering
\includegraphics[width=.7\textwidth]{graphics/bickley_jet/lambda_lcs_convergence_400}
\includegraphics[width=.7\textwidth]{graphics/bickley_jet/lambda_lcs_convergence_500}
\includegraphics[width=.7\textwidth]{graphics/bickley_jet/lambda_lcs_convergence_600}
\caption{Bickley jet closed lambda-line at varying resolutions}
\label{fig:Bickley jet lambda lcs convergence}
\end{figure}

@@

% Lambda-line LCSs
% Define Poincare sections; ...
poincareSection = struct('endPosition',{},'numPoints',{},'orbitMaxLength',{});
...
poincareSection(1).endPosition = [3.3,-32.1;3.7,-31.6];
poincareSection(2).endPosition = [1.3,-30.9;1.9,-31.1];
